Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments

### 2016-05-31 paper.tex
Minor fixes. Author affiliations.

\author[1]{Daniel C. Jones}
\author[2]{Kavitha T. Kuppusamy}
\author[3]{Nathan J. Palpant}
\author[4]{Xinxia Peng}
\author[4]{Michael G. Katze}
\author[5]{Charles E. Murry}
\author[6]{Hannele Ruohola-Baker}
\author[1,7,8]{Walter L. Ruzzo}

@@

\affil[1]{Department of Computer Science and Engineering, University of Washington}
\affil[2]{XXX}
\affil[3]{XXX}
\affil[4]{Department of Microbiology, University of Washington}
\affil[5]{Department of Pathology, University of Washington}
\affil[6]{Department of Biochemistry, University of Washington}
\affil[7]{Department of Genome Sciences, University of Washington}
\affil[8]{Fred Hutchinson Cancer Research Center}

@@

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto.
% TODO: see supplementary section XXX (how reads were aligned)

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between
samples sequenced on two flowcells each at five sites, from centrally prepared
libraries. Flowcells are numbered arbitrarily 1 or 2 and sequencing sites are
abbreviated with three letter codes: Australian Genome Research Facility (AGR),
Beijing Genome Institute (BGI), Cornell University (CNL), Mayo Clinic (MAY), and
Novartis (NVS). \textbf{b} The absolute change in correlation induced by
enabling bias correction for methods that support it. For clarity this plot
excludes points for BitSeq estimates of "MAY 2", as bias correction has an
extremely detrimental effect on these. Mean improvement in correlation was 0.008
with Salmon, 0.007 with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto,
0.002 with eXpress, and -0.164 with BitSeq.
}

@@

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each consisting of a single lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement between estimates from these
samples (Figure \ref{fig:batcheffects}a).

@@

\begin{table}
\label{table:rlsim}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & \textbf{0.919}
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{
Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\end{table}

@@

Isolator shows the highest correlation, but only when sequencing depth
exceeds a few million reads (below which would be unusually shallow
sequencing for an RNA-Seq experiment) where it overtakes eXpress by a small
margin. The outliers in the this test are BitSeq and RSEM/PM. The former
shows extremely high correlation with small numbers of reads, but
contrary to all other methods, correlation rapidly decreases with deeper
sequencing after 1 million reads. When posterior mean estimates are
generated from a model using a uninformative prior, as is the case with
BitSeq, transcripts with few or no reads may produce estimates that are
highly influenced by transcript length, which plausibly explains this
phenomenon. Yet, RSEM/PM does increase in correlation monotonically with
sequencing depth, though the correlation remains very low. Even using
all 11 million reads, RSEM/PM had a correlation of only 0.567.

@@

Splicing has been known to play an important role in devoloping hearts. Specific
splicing changes during maturation have been shown to be conserved across
chicken and mouse, and influenced by expression of CELF and MBNL
\cite{Kalsotra:2008ke}. Recently, stress-induced switching between two mutually
exclusive isoforms of KHK has been shown to be driver of pathological cardiac
hypertrophy \cite{Mirtschink:2015iq}.

@@

We instead searched for differential splicing events by computing the
probability of ``monotonic splicing``, which we defined as a splicing change
that that occurs is consistently observed as occuring in the same direction
between immature (H7 cells at day 20 and at day 30 infected with an empty
vector) and mature samples (H7 1 Year, adult heart, and fetal atrium and
ventricle). To compute the posterior probability of monotonic splicing, we
simply count the number of samples from the model in which this pattern is
observed, and divide by the total number of samples generated (350, in this
case).

This approach identified 272 probable (posterior probability > 0.5) differential
splicing events. We chose seven high probability candidates (ACOT9, IMMT, LTBP4,
OBSL1, OSBPL1A, PALM, and TNNT2) to validate with qPCR (Supplementary Section
\ref{cardioqpcr}), by measuring the expression of specific exons predicted a
change in splicing. Five of these candidates (ACOT9, LTBP4, OBSL1, OSBPL1A, and
TNNT2) showed a clear gradient between H7 day20, fetal, and adult tissue.

@@

\subsection{Case study: novel mechanisms of cell fate specification identified
by gene splice variants}

\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{palpant/figure.pdf}
\end{center}
\caption{
\textbf{a}
Hierarchical clustering of the dominant splice variant for 84
differentially spliced genes. Blue arrows indicate genes with known function in
endothelial or heart development. Red arrows indicate genes in the class of
receptors that bind VEGF.
\textbf{b}
Two dimensional principle component analysis in mesodermal progenitor
populations showing discrete separation of populations entirely based on
differences in splice variants. 
\textbf{c}
Analysis of gene ontology showing enriched categories found in 84 differentially
spliced genes. (d-e) Heat map analysis 
\textbf{d}
Heat map showing expression of annotated isoforms of NRP1.
\textbf{e}
Read coverage over isoforms of NRP1.
\textbf{f}
Schematic diagram of protein domains encoded by splice variant variants of NRP1.}
\label{fig:mesodermal}
\end{figure*}

@@

To assess the suitability of Isolator we have compared it to large number of
alternative methods using a wide variety of benchmarks, with a particular focus
on estimation consistency or stability. Data generated by the SEQC is ideally
suited to examine the question consistency, and indeed our data agrees with
their results. The SEQC compared Cufflinks and BitSeq and found BitSeq to
produce more consistent estimates. Here we control for alignment method by using
the same alignments for every method (excluding the alignment-free methods
Salmon, Sailfish, and Kallisto), and compare wider variety methods. In this
broader context it becomes clear that this separation is a consequence of the
choice in estimators. Unmoderated maximum likelihood methods, though often
accurate, frequently demonstrate lower consistency than methods that use MCMC
sampling and report posterior means or medians. This should be unsurprising.
Because of the ambiguity of sequencing data, the likelihood functions for
relative expression are often near plateaus. Maximum likelihood estimates are
thus are precarious compared to posterior mean estimates. Neither estimate is
inherently more correct, and robust analysis should make use of the full
posterior distribution, but we argue that posterior mean or median estimates
are more useful in practice because of this property.

Most methods that report posterior mean estimates show seemingly poor accuracy.
Much of the cause of the seemingly reduced accuracy of the MCMC based models is
driven by low-expression transcripts. A transcript with no reads has a maximum
likelihood expression of zero, but likelihood is positive over a range of
non-zero values, so the posterior mean must necessarily also be positive. With
a flat, or ``non-informative'' prior expression values for these low-expression
transcripts are often quite high and highly determined by transcript length.
Isolator avoids this problem by using a model with informative priors: namely
encoding assumptions that most transcripts have low expression and that
expression and splicing tend to be similar between samples. In this way, it is
able to avoid what seems like a tradeoff and produce estimates that are both
consistent and
accurate.

Fully Bayesian methods like the one presented here possess a degree of
subjectivity that sometimes give researchers pause. Although we have
found the results from Isolator to be insensitive to precise values
given to hyperparameters, they are chosen in advance and with a degree
of arbitrariness. What is often ignored is that the alternative, methods
based on unregularized maximum likelihood estimates, in practice often
necessitate a more insidious form of subjectivity: contriving cutoffs to
filter out unreliable point estimates of low expression genes or
transcripts. While the priors in Bayesian methods are purely explicit,
the means by which data was filtered during an analysis is often a form
of off-the-books subjectivity--critical to the results, yet unmentioned
or only alluded to in manuscripts.

@@

To model RNA-Seq experiments we use a hierarchical model consisting of
three levels: samples or replicates, conditions, and the experiment. Any
number of conditions, and any number of replicates within those
conditions. We model condition-wise and experiment-wise expression for
each transcript using a Gamma distribution, parameterized by mean and
shape. Shape, which capture biological variability, are shared across
conditions.

@@

@@

The isoform quantification or deconvolution problem is most easily
thought of as an additive mixture model in which the component
distributions (the isoforms) are known, but the mixture coefficients
(isoform abundances) must be inferred. The standard approach to this
problem is ether expectation maximization (EM) if the maximum likelihood
solution is desired, or Gibbs sampling if, as we are, estimating the
posterior distribution. We have taken a different approach and instead
rely on slice sampling.

The trade-offs between slice sampling and Gibbs sampling for the isoform
quantification problem are subtle. Subsequent samples drawn from a slice
sampler are generally less autocorrelated than those drawn from a Gibbs
sampler. As a result, fewer samples need to be generated to adequately
explore a distribution. However, each of these samples is more
computationally expensive to compute, so one might compensate for
autocorrelation in a Gibbs sampler by drawing more samples in less time.
It is not obvious then that one approach is inherently more efficient
than the other.

Because the likelihood function in this problem is simple, and we sample
from the posterior probability directly, without introducing latent
variables, we are afforded some optimization opportunities not available
in a Gibbs sampler. Specifically, we implement the likelihood function
using SIMD (single instruction, multiple data) instructions, which are
available nearly all CPUs from the last decade. This, in combination
with some numerical approximations, allow us to compute the likelihood
function over an order of magnitude faster on conventional hardware,
than a more straightforward implementation. Run time for Isolator is generally
significantly less than that of other sampling based implementations, though
alignment-free, maximum-likelihood methods like Salmon and Kallisto remain the
most efficient methods (Supplementary Figure \ref{fig:runtime})

@@

Perhaps the most prominent source of nonuniformity is sequence specific
bias present at fragment ends. A probable source of this bias is cDNA
synthesis by random priming, which is part of popular RNA-Seq protocols
such as Illumina's TruSeq kit. We have previously published a model that
successfully accounts for effect by training a sparse graphical model
\cite{Jones:2012hg}.

Beyond this random priming bias, there is residual fragment GC-bias,
which may be an artifact of PCR amplification. We observe that fragments
with very high or very low GC-content are less likely to be observed
than expected, even after accounting for priming bias. We model this by
binning fragments by GC content, computing expected and observed
frequencies, and adjusting by their ratio.

Selection of polyadenylated transcripts is a common step mRNA-Seq used
to avoid sequencing introns, partially degraded transcripts, and
ribosomal RNA. This can cause enrichment of reads at the 3' end of
transcripts if only partial transcripts are captured. To model the
effect we fit a one parameter model in which a transcript is truncated
at any position with probability $p$. The probability of observing a
fragment at a position $k$ of length $\ell$, counting from the 3' end,
is then proportional to the probability that the transcript was
\emph{not} truncated before $k + \ell$, which is $(1-p)^{k + \ell}$.
When fit, $p$ is typically quite small, on the order of $10^{-5}$, in
which case this correction has little effect on transcripts shorter that
several kilobases.

Lastly, subtle implications arise from random fragmentation steps in many
protocols. Existing statistical models assume fragments are sampled uniformly
at random from a transcript. However, this does not exactly match the
implications of random fragmentation in which, under an ideal model,
\emph{breakpoints} rather than fragments are introduced uniformly at random. For
a fragment to be observed it must pass size selection and have fallen between
two breakpoints, or one breakpoint and the end of the transcript. Since the ends
of a transcript act as fixed breakpoints, the result is some enrichment of
fragments at either end of the transcript. We compute this enrichment exactly
for various transcript length, and use interpolation to approximate th effect.

>>>>

\author[1]{Daniel C. Jones}
\author[2,6]{Kavitha T. Kuppusamy}
\author[3]{Nathan J. Palpant}
\author[4]{Xinxia Peng}
\author[4]{Michael G. Katze}
\author[2,5]{Charles E. Murry}
\author[2,6]{Hannele Ruohola-Baker}
\author[1,7,8]{Walter L. Ruzzo}

@@

\affil[1]{Department of Computer Science and Engineering, University of Washington}
\affil[2]{Institute for Stem Cell and Regenerative Medicine}
%\affil[3]{Institute for Molecular Bioscience, The University of Queensland. Brisbane. Australia}
\affil[3]{Institute for Molecular Bioscience, The University of Queensland}
\affil[4]{Department of Microbiology, University of Washington}
\affil[5]{Department of Pathology, University of Washington}
\affil[6]{Department of Biochemistry, University of Washington}
\affil[7]{Department of Genome Sciences, University of Washington}
\affil[8]{Fred Hutchinson Cancer Research Center}

@@

Regardless, splicing has been repeatedly shown to be a essential regulatory
mechanism, and particularly important in development and differentiation
\cite{Kalsotra:2011hv}. We examine alternative splicing in heart development in
two contexts in Sections \ref{finding-monotonic-differential-splicing} and
\ref{novel-mechanisms-of-cell-fate}.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto (Supplementary Section \ref{analysis}).

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between
samples sequenced on two flowcells each at five sites, from centrally prepared
libraries. Flowcells are numbered arbitrarily 1 or 2 and sequencing sites are
abbreviated with three letter codes: Australian Genome Research Facility (AGR),
Beijing Genome Institute (BGI), Cornell University (CNL), Mayo Clinic (MAY), and
Novartis (NVS). Median proportionality correlation is listed below each heatmap.
\textbf{b} The absolute change in correlation induced by enabling bias
correction for methods that support it. For clarity this plot excludes points
for BitSeq estimates of "MAY 2", as bias correction has an extremely detrimental
effect on these. Mean improvement in correlation was 0.008 with Salmon, 0.007
with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto, 0.002 with eXpress,
and -0.164 with BitSeq.
}

@@

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each consisting of a single lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement between estimates from these
samples (Figure \ref{fig:batcheffects}a). In the interest of a more fair
comparison, Isolator was run on each sample in isolation without sharing
any information.

@@

\begin{table}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & \textbf{0.919}
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{
Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\label{table:rlsim}
\end{table}

@@

Isolator shows the highest correlation, but only when sequencing depth
exceeds a few million reads (below which would be unusually shallow
sequencing for an RNA-Seq experiment) where it overtakes eXpress by a small
margin. The outliers in the this test are BitSeq and RSEM/PM. The former
shows extremely high correlation with small numbers of reads, but
contrary to all other methods, correlation rapidly decreases with
sequencing deeper than 1 million reads. When posterior mean estimates are
generated from a model using a uninformative prior, as is the case with
BitSeq, transcripts with few or no reads may produce estimates that are
highly influenced by transcript length, which plausibly explains this
phenomenon. Yet, RSEM/PM does increase in correlation monotonically with
sequencing depth, though the correlation remains very low. Even using
all 11 million reads, RSEM/PM had a correlation of only 0.567 (data not shown).

@@

Splicing has been known to play an important role in developing hearts. Specific
splicing changes during maturation have been shown to be conserved across
chicken and mouse, and influenced by expression of CELF and MBNL
\cite{Kalsotra:2008ke}. Recently, stress-induced switching between two mutually
exclusive isoforms of KHK has been shown to be driver of pathological cardiac
hypertrophy \cite{Mirtschink:2015iq}.

@@

We instead searched for differential splicing events by computing the
probability of ``monotonic splicing``, which we defined as a splicing change
that that is consistently observed to occur in the same direction
between immature (H7 cells at day 20 and at day 30 infected with an empty
vector) and mature samples (H7 1 Year, adult heart, and fetal atrium and
ventricle). To compute the posterior probability of monotonic splicing, we
simply count the number of samples from the model in which this pattern is
observed, and divide by the total number of samples generated (350, in this
case).

This approach identified 272 probable (posterior probability > 0.5) differential
splicing events. We chose seven high probability candidates (ACOT9, IMMT, LTBP4,
OBSL1, OSBPL1A, PALM, and TNNT2) to validate with qPCR (Supplementary Section
\ref{cardioqpcr}), by measuring the expression of specific exons predicted to
change in splicing. Five of these candidates (ACOT9, LTBP4, OBSL1, OSBPL1A, and
TNNT2) showed a clear gradient between H7 day20, fetal, and adult tissue.

@@

\subsection{Case study: novel mechanisms of cell fate specification identified
by gene splice variants}\label{novel-mechanisms-of-cell-fate}

\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{palpant/figure.pdf}
\end{center}
\caption{
\textbf{a}
Hierarchical clustering of the dominant splice variant for 84
differentially spliced genes. Blue arrows indicate genes with known function in
endothelial or heart development. Red arrows indicate genes in the class of
receptors that bind VEGF.
\textbf{b}
Two dimensional principle component analysis in mesodermal progenitor
populations showing discrete separation of populations entirely based on
differences in splice variants.
\textbf{c}
Analysis of gene ontology showing enriched categories found in 84 differentially
spliced genes.
\textbf{d}
Heat map showing expression of annotated isoforms of NRP1.
\textbf{e}
Read coverage over isoforms of NRP1.
\textbf{f}
Schematic diagram of protein domains encoded by splice variant variants of NRP1.}
\label{fig:mesodermal}
\end{figure*}

@@

To assess the suitability of Isolator we have compared it to large number of
alternative methods using a wide variety of benchmarks, with a particular focus
on estimation consistency or stability. Data generated by the SEQC is ideally
suited to examine the question consistency, and indeed our data agrees with
their results. The SEQC compared Cufflinks and BitSeq and found BitSeq to
produce more consistent estimates. Here we control for alignment method by using
the same alignments for every method (excluding the alignment-free methods
Salmon, Sailfish, and Kallisto), and compare a wider variety methods. In this
broader context it becomes clear that this separation is a consequence of the
choice in estimators. Unmoderated maximum likelihood methods, though often
accurate, frequently demonstrate lower consistency than methods that use MCMC
sampling and report posterior means or medians. This is underappreciated but
should be unsurprising.  Because of the ambiguity of sequencing data, the
likelihood functions for relative expression are often near plateaus. Maximum
likelihood estimates are thus precarious compared to posterior mean
estimates. Neither estimate is inherently more correct, and robust analysis
should make use of the full posterior distribution, but we argue that posterior
mean or median estimates are more useful in practice because of their greater
stability.

Excluding Isolator, methods that report posterior mean estimates show seemingly
poor accuracy. This is largely driven by low-expression transcripts. A
transcript with no reads has a maximum likelihood expression of zero, but its
likelihood is positive over a range of non-zero values, so the posterior mean
must necessarily also be positive. With a flat, or ``non-informative,'' prior,
expression values for these low-expression transcripts are often quite high and
highly determined by transcript length.  Isolator avoids this problem by using a
model with informative priors: namely encoding assumptions that most transcripts
have low expression and that expression and splicing tend to be similar between
samples. In this way, it is able to avoid what seems like a tradeoff and
produces estimates that are both consistent and accurate.

Fully Bayesian methods like the one presented here possess a degree of
subjectivity that sometimes give researchers pause. Although we have found the
results from Isolator to be insensitive to precise values given to
hyperparameters, they are chosen in advance and with a degree of arbitrariness.
What is often ignored is that the alternative, methods based on unregularized
maximum likelihood estimates, in practice often necessitate a more insidious
form of subjectivity: contriving cutoffs to filter out unreliable point
estimates of low expression genes or transcripts. While the priors in Bayesian
methods are purely explicit, ad hoc filtering of unregularized estimates can
often be a form of off-the-books subjectivity--critical to the results, yet
unmentioned or only alluded to in manuscripts.

In this paper we have described a method that effectively shares information
between samples in RNA-Seq experiments while rigorously quantifying uncertainty,
leading to more accurate results across a wide range of benchmarks. We believe
this work indicates a natural direction towards more reliable and powerful
studies of transcriptomics, which future work can follow by scaling this
methodology to larger numbers of samples, and accounting for more forms of
uncertainty and technical effects.

@@

Input to Isolator consists of gene annotations in GTF/GFF format, one
or more sorted BAM \cite{Li:2009ka} file, and optionally a reference sequence,
which if provided is used in bias correction.

To model RNA-Seq experiments we use a hierarchical model consisting of
three levels: samples or replicates, conditions, and the experiment, with any
number of conditions, and any number of replicates within those
conditions. We model condition-wise and experiment-wise expression for
each transcript using a Gamma distribution, parameterized by mean and
shape. Shape, which capture biological variability, are shared across
conditions.

@@

Parameters are updated in sequence using a Gibbs sampling approach. The output
generated by Isolator consists of each sample of every parameter serialized in
an HDF5-based format, along with basic diagnostic information.  Point estimates,
summary statistics, and posterior probabilities can then be calculated from this
output using a secondary command (\texttt{isolator summarize}).

@@

The isoform quantification or deconvolution problem is most easily
thought of as an additive mixture model in which the component
distributions (the isoforms) are known, but the mixture coefficients
(isoform abundances) must be inferred. The standard approach to this
problem is either expectation maximization (EM) if the maximum likelihood
solution is desired, or Gibbs sampling if, as we are, estimating the
posterior distribution. We have taken a different approach and instead
rely on slice sampling \cite{Neal:2000uc}.

The trade-offs between slice sampling and Gibbs sampling for the isoform
quantification problem are subtle. Subsequent samples drawn from a slice
sampler are generally less autocorrelated than those drawn from a Gibbs
sampler. As a result, fewer samples need to be generated to adequately
explore a distribution. However, each of these samples is more
computationally expensive to compute, so one might compensate for
autocorrelation in a Gibbs sampler by drawing more samples in less time.
It is not obvious, then, that one approach is inherently more efficient
than the other.

Because the likelihood function in this problem is simple, and we sample from
the posterior probability directly, without introducing latent variables, we are
afforded some optimization opportunities not available in a Gibbs sampler.
Specifically, we implement the likelihood function using SIMD (single
instruction, multiple data) instructions, which are available nearly all CPUs
from the last decade. This, in combination with some numerical approximations,
allow us to compute the likelihood function over an order of magnitude faster,
than a more straightforward implementation. Run time for Isolator is generally
significantly less than that of other sampling based implementations, though
alignment-free, maximum-likelihood methods like Salmon and Kallisto remain the
most efficient methods (Supplementary Figure \ref{fig:runtime})

@@

Perhaps the most prominent source of nonuniformity is sequence specific bias
present at fragment ends. A probable contributor of this bias is cDNA synthesis
by random priming, which is part of popular RNA-Seq protocols such as Illumina's
TruSeq kit. We have previously published a model that successfully accounts for
such effects by training a sparse graphical model \cite{Jones:2012hg}.

Beyond this sequence context bias, there is residual fragment GC-bias,
which may be an artifact of PCR amplification. We observe that fragments
with very high or very low GC-content are less likely to be sequenced
than expected, even after accounting for sequence bias. We model this by
binning fragments by GC content, computing expected and observed
frequencies, and adjusting by their ratio.

Selection of polyadenylated transcripts is a common step mRNA-Seq, used to avoid
sequencing introns, partially degraded transcripts, and ribosomal RNA. This can
cause enrichment of reads at the 3' end of transcripts if only partial
transcripts are captured. To model the effect we fit a one parameter model in
which a transcript is truncated at any position with probability $p$. The
probability of observing a fragment ending at a position $k$, counting from the
3' end, is then proportional to the probability that the transcript was
\emph{not} truncated before $k$, which is $(1-p)^{k}$.  When fit, $p$ is
typically quite small, on the order of $10^{-5}$, in which case this correction
has little effect on transcripts shorter that several kilobases.

Lastly, subtle implications arise from random fragmentation steps in many
protocols. Existing statistical models assume fragments are sampled uniformly
at random from a transcript. However, this does not exactly match the
implications of random fragmentation in which, under an ideal model,
\emph{breakpoints} rather than fragments are introduced uniformly at random. For
a fragment to be observed it must pass size selection and have fallen between
two breakpoints, or one breakpoint and the end of the transcript. Since the ends
of a transcript act as fixed breakpoints, the result is some enrichment of
fragments at either end of the transcript. We compute this enrichment exactly
for various transcript length, and use interpolation to approximate the effect.
